---
title: "Grant Sharma"
author: "Erik Parker and Lijiang Gou for the Biostatistics Consulting Center (BCC)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 2
    code_folding: hide
    df_print: kable
---

```{=html}
<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(plyr)
library(tidyverse)
library(magrittr)
library(kableExtra)
library(ggrepel)
library(patchwork)

pcr.data <- read_csv("../Data/Analysis data/qPCRmice cleaned analysis data 2023-12-13.csv") %>%
  mutate(target = factor(target),
         Treatment = factor(Treatment),
         sex = factor(sex)) %>%
  rename(treatment = Treatment)

```

# Number of significant fold changes, 0% nicotine vs. control, by sex {.tabset}


## Males

```{r}

test.results <- NULL
# make new empty variable that will hold the results

males <- pcr.data %>%
  filter(sex == "m" & (treatment == "control" | treatment == "ecig0"))
# filter data to just include males and the treatments that we know

for (i in levels(pcr.data$target)) {
  # initiate loop over the levels of the `target` variable to fit one model per target
  
  data.loop = subset(males, target == i)
  # subset the data to just have one target per loop
  
  data.loop <- data.loop %>% mutate(treatment = factor(treatment, levels = c("control",
                                                                     "ecig0")))
  # relevel the dataset so that the negative control is the reference level. Gets us more interpretable fold change cofficients 
  
  model <- lm(log_fold_change ~ treatment, data = data.loop)
  # fit the model (here the equivalent of a t-test since there are only two levels to the treatment)
  
  anova <- car::Anova(model)
  # fit the anova on the model to get a p-value
  
  # save p value from model for each target
  test.results.list <- data.frame(target = i,
                               p.val = anova$`Pr(>F)`[1],
                               coef = coef(model)[[2]])

  test.results <- rbind(test.results, test.results.list)
  # save the results from each loop into the dataframe set up at the beginning to hold them

}

```

### {.tabset}

#### N targets significantly different

```{r}

kable(test.results %>%
        summarise(sum(p.val < 0.05)), col.names = "N targets significantly different (p < 0.05)") %>%
  kableExtra::kable_styling(full_width = FALSE)
# calculate the number of targets with p-values < 0.05

```

#### Significantly different targets (w/ pvalues and average log FCs)

```{r}

kable(test.results %>% filter(p.val < 0.05), col.names = c("target", "p.value", "log fold change"), caption = "Significantly different targets", digits = 3) %>%
  kableExtra::kable_styling(full_width = FALSE)
# list the targets falling below the 0.05 threshold

```

#### "Volcano plot"

I hand-made the rt-qPCR equivalent of an RNA-SEQ volcano plot with the natural log fold change plotted against the -log base 10 p-value for each target's control vs. experimental comparison.  
This simply serves as a way to visually represent the tables given above. But they also look cute!  

```{r, fig.width=8, fig.height=6}

test.results <- test.results %>% mutate(color = ifelse(p.val >= 0.05, "Not significant", NA),
                                        color = ifelse(p.val < 0.05 & coef < 0, "Decrease", color),
                                        color = ifelse(p.val < 0.05 & coef >= 0, "Increase", color),
                                        color = as.factor(color))
# make new variable to color points based on p-value

ggplot(test.results, aes(x = coef, y = -log10(p.val), color = color)) + # plot p.val on y axis, log(FC) on x
    geom_point() + # add points colored by p-value level
    geom_abline(intercept = -log10(0.05), slope = 0) + # add a line at equivalent of p = 0.05
    geom_label_repel(aes(label = ifelse(p.val < 0.05, as.character(target), '')), hjust = 0, vjust = 0) + #label targets p <0.05
    labs(x = "log2(fold change)", y = "-log10(p value)", color = "Directionality") + # add plot labels
    theme_bw()

```


## Females

```{r}

test.results <- NULL

females <- pcr.data %>%
  filter(sex == "f" & (treatment == "control" | treatment == "ecig0"))

for (i in levels(pcr.data$target)) {
  
  data.loop = subset(females, target == i)
  
  data.loop <- data.loop %>% mutate(treatment = factor(treatment, levels = c("control",
                                                                               "ecig0")))
  
  model <- lm(log_fold_change ~ treatment, data = data.loop)
  
  anova <- car::Anova(model)
  
  # save p value from model for each target
  test.results.list <- data.frame(target = i,
                               p.val = anova$`Pr(>F)`[1],
                               coef = coef(model)[[2]])

  test.results <- rbind(test.results, test.results.list)

}

```

### {.tabset}

#### N targets significantly different

```{r}

kable(test.results %>% summarise(sum(p.val < 0.05)), col.names = "N targets significantly different (p < 0.05)") %>%
  kableExtra::kable_styling(full_width = FALSE)

```

#### Significantly different targets (w/ pvalues and average log FCs)

```{r}

kable(test.results %>% filter(p.val < 0.05), col.names = c("target", "p.value", "log fold change"), caption = "Significantly different targets", digits = 3) %>%
  kableExtra::kable_styling(full_width = FALSE)

```

#### "Volcano plot"

```{r, fig.width=8, fig.height=6}

test.results <- test.results %>% mutate(color = ifelse(p.val >= 0.05, "Not significant", NA),
                                        color = ifelse(p.val < 0.05 & coef < 0, "Decrease", color),
                                        color = ifelse(p.val < 0.05 & coef >= 0, "Increase", color),
                                        color = as.factor(color))

ggplot(test.results, aes(x = coef, y = -log10(p.val), color = color)) + 
    geom_point() + 
    geom_abline(intercept = -log10(0.05), slope = 0) + 
    geom_label_repel(aes(label = ifelse(p.val < 0.05, as.character(target), '')), hjust = 0, vjust = 0) +
    labs(x = "log2(fold change)", y = "-log10(p value)", color = "Directionality") +
    theme_bw()

```



# Number of significant fold changes, 3% nicotine vs. control, by sex {.tabset}


## Males

```{r}

test.results <- NULL
# make new empty variable that will hold the results

males <- pcr.data %>%
  filter(sex == "m" & (treatment == "control" | treatment == "ecig3"))
# filter data to just include males and the treatments we want here

for (i in levels(pcr.data$target)) {
  # initiate loop over the levels of the `target` variable to fit one model per target
  
  data.loop = subset(males, target == i)
  # subset the data to just have one target per loop
  
  data.loop <- data.loop %>% mutate(treatment = factor(treatment, levels = c("control",
                                                                     "ecig3")))
  # relevel the dataset so that the negative control is the reference level. Gets us more interpretable fold change cofficients 
  
  model <- lm(log_fold_change ~ treatment, data = data.loop)
  # fit the model (here the equivalent of a t-test since there are only two levels to the treatment)
  
  anova <- car::Anova(model)
  # fit the anova on the model to get a p-value
  
  # save p value from model for each target
  test.results.list <- data.frame(target = i,
                               p.val = anova$`Pr(>F)`[1],
                               coef = coef(model)[[2]])

  test.results <- rbind(test.results, test.results.list)
  # save the results from each loop into the dataframe set up at the beginning to hold them

}

```

### {.tabset}

#### N targets significantly different

```{r}

kable(test.results %>%
        summarise(sum(p.val < 0.05)), col.names = "N targets significantly different (p < 0.05)") %>%
  kableExtra::kable_styling(full_width = FALSE)
# calculate the number of targets with p-values < 0.05

```

#### Significantly different targets (w/ pvalues and average log FCs)

```{r}

kable(test.results %>% filter(p.val < 0.05), col.names = c("target", "p.value", "log fold change"), caption = "Significantly different targets", digits = 3) %>%
  kableExtra::kable_styling(full_width = FALSE)
# list the targets falling below the 0.05 threshold

```

#### "Volcano plot"

I hand-made the rt-qPCR equivalent of an RNA-SEQ volcano plot with the natural log fold change plotted against the -log base 10 p-value for each target's control vs. experimental comparison.  
This simply serves as a way to visually represent the tables given above. But they also look cute!  

```{r, fig.width=8, fig.height=6}

test.results <- test.results %>% mutate(color = ifelse(p.val >= 0.05, "Not significant", NA),
                                        color = ifelse(p.val < 0.05 & coef < 0, "Decrease", color),
                                        color = ifelse(p.val < 0.05 & coef >= 0, "Increase", color),
                                        color = as.factor(color))
# make new variable to color points based on p-value

ggplot(test.results, aes(x = coef, y = -log10(p.val), color = color)) + # plot p.val on y axis, log(FC) on x
    geom_point() + # add points colored by p-value level
    geom_abline(intercept = -log10(0.05), slope = 0) + # add a line at equivalent of p = 0.05
    geom_label_repel(aes(label = ifelse(p.val < 0.05, as.character(target), '')), hjust = 0, vjust = 0) + #label targets p <0.05
    labs(x = "log2(fold change)", y = "-log10(p value)", color = "Directionality") + # add plot labels
    theme_bw()

```


## Females

```{r}

test.results <- NULL

females <- pcr.data %>%
  filter(sex == "f" & (treatment == "control" | treatment == "ecig3"))

for (i in levels(pcr.data$target)) {
  
  data.loop = subset(females, target == i)
  
  data.loop <- data.loop %>% mutate(treatment = factor(treatment, levels = c("control",
                                                                               "ecig3")))
  
  model <- lm(log_fold_change ~ treatment, data = data.loop)
  
  anova <- car::Anova(model)
  
  # save p value from model for each target
  test.results.list <- data.frame(target = i,
                               p.val = anova$`Pr(>F)`[1],
                               coef = coef(model)[[2]])

  test.results <- rbind(test.results, test.results.list)

}

```

### {.tabset}

#### N targets significantly different

```{r}

kable(test.results %>% summarise(sum(p.val < 0.05)), col.names = "N targets significantly different (p < 0.05)") %>%
  kableExtra::kable_styling(full_width = FALSE)

```

#### Significantly different targets (w/ pvalues and average log FCs)

```{r}

kable(test.results %>% filter(p.val < 0.05), col.names = c("target", "p.value", "log fold change"), caption = "Significantly different targets", digits = 3) %>%
  kableExtra::kable_styling(full_width = FALSE)

```

#### "Volcano plot"

```{r, fig.width=8, fig.height=6}

test.results <- test.results %>% mutate(color = ifelse(p.val >= 0.05, "Not significant", NA),
                                        color = ifelse(p.val < 0.05 & coef < 0, "Decrease", color),
                                        color = ifelse(p.val < 0.05 & coef >= 0, "Increase", color),
                                        color = as.factor(color))

ggplot(test.results, aes(x = coef, y = -log10(p.val), color = color)) + 
    geom_point() + 
    geom_abline(intercept = -log10(0.05), slope = 0) + 
    geom_label_repel(aes(label = ifelse(p.val < 0.05, as.character(target), '')), hjust = 0, vjust = 0) +
    labs(x = "log2(fold change)", y = "-log10(p value)", color = "Directionality") +
    theme_bw()

```




# Number of significant fold changes, 6% nicotine vs. control, by sex {.tabset}


## Males

```{r}

test.results <- NULL
# make new empty variable that will hold the results

males <- pcr.data %>%
  filter(sex == "m" & (treatment == "control" | treatment == "ecig6"))
# filter data to just include males and the treatments that we know

for (i in levels(pcr.data$target)) {
  # initiate loop over the levels of the `target` variable to fit one model per target
  
  data.loop = subset(males, target == i)
  # subset the data to just have one target per loop
  
  data.loop <- data.loop %>% mutate(treatment = factor(treatment, levels = c("control",
                                                                     "ecig6")))
  # relevel the dataset so that the negative control is the reference level. Gets us more interpretable fold change cofficients 
  
  model <- lm(log_fold_change ~ treatment, data = data.loop)
  # fit the model (here the equivalent of a t-test since there are only two levels to the treatment)
  
  anova <- car::Anova(model)
  # fit the anova on the model to get a p-value
  
  # save p value from model for each target
  test.results.list <- data.frame(target = i,
                               p.val = anova$`Pr(>F)`[1],
                               coef = coef(model)[[2]])

  test.results <- rbind(test.results, test.results.list)
  # save the results from each loop into the dataframe set up at the beginning to hold them

}

```

### {.tabset}

#### N targets significantly different

```{r}

kable(test.results %>%
        summarise(sum(p.val < 0.05)), col.names = "N targets significantly different (p < 0.05)") %>%
  kableExtra::kable_styling(full_width = FALSE)
# calculate the number of targets with p-values < 0.05

```

#### Significantly different targets (w/ pvalues and average log FCs)

```{r}

kable(test.results %>% filter(p.val < 0.05), col.names = c("target", "p.value", "log fold change"), caption = "Significantly different targets", digits = 3) %>%
  kableExtra::kable_styling(full_width = FALSE)
# list the targets falling below the 0.05 threshold

```

#### "Volcano plot"

I hand-made the rt-qPCR equivalent of an RNA-SEQ volcano plot with the natural log fold change plotted against the -log base 10 p-value for each target's control vs. experimental comparison.  
This simply serves as a way to visually represent the tables given above. But they also look cute!  

```{r, fig.width=8, fig.height=6}

test.results <- test.results %>% mutate(color = ifelse(p.val >= 0.05, "Not significant", NA),
                                        color = ifelse(p.val < 0.05 & coef < 0, "Decrease", color),
                                        color = ifelse(p.val < 0.05 & coef >= 0, "Increase", color),
                                        color = as.factor(color))
# make new variable to color points based on p-value

ggplot(test.results, aes(x = coef, y = -log10(p.val), color = color)) + # plot p.val on y axis, log(FC) on x
    geom_point() + # add points colored by p-value level
    geom_abline(intercept = -log10(0.05), slope = 0) + # add a line at equivalent of p = 0.05
    geom_label_repel(aes(label = ifelse(p.val < 0.05, as.character(target), '')), hjust = 0, vjust = 0) + #label targets p <0.05
    labs(x = "log2(fold change)", y = "-log10(p value)", color = "Directionality") + # add plot labels
    theme_bw()

```


## Females

```{r}

test.results <- NULL

females <- pcr.data %>%
  filter(sex == "f" & (treatment == "control" | treatment == "ecig6"))

for (i in levels(pcr.data$target)) {
  
  data.loop = subset(females, target == i)
  
  data.loop <- data.loop %>% mutate(treatment = factor(treatment, levels = c("control",
                                                                               "ecig6")))
  
  model <- lm(log_fold_change ~ treatment, data = data.loop)
  
  anova <- car::Anova(model)
  
  # save p value from model for each target
  test.results.list <- data.frame(target = i,
                               p.val = anova$`Pr(>F)`[1],
                               coef = coef(model)[[2]])

  test.results <- rbind(test.results, test.results.list)

}

```

### {.tabset}

#### N targets significantly different

```{r}

kable(test.results %>% summarise(sum(p.val < 0.05)), col.names = "N targets significantly different (p < 0.05)") %>%
  kableExtra::kable_styling(full_width = FALSE)

```

#### Significantly different targets (w/ pvalues and average log FCs)

```{r}

kable(test.results %>% filter(p.val < 0.05), col.names = c("target", "p.value", "log fold change"), caption = "Significantly different targets", digits = 3) %>%
  kableExtra::kable_styling(full_width = FALSE)

```

#### "Volcano plot"

```{r, fig.width=8, fig.height=6}

test.results <- test.results %>% mutate(color = ifelse(p.val >= 0.05, "Not significant", NA),
                                        color = ifelse(p.val < 0.05 & coef < 0, "Decrease", color),
                                        color = ifelse(p.val < 0.05 & coef >= 0, "Increase", color),
                                        color = as.factor(color))

ggplot(test.results, aes(x = coef, y = -log10(p.val), color = color)) + 
    geom_point() + 
    geom_abline(intercept = -log10(0.05), slope = 0) + 
    geom_label_repel(aes(label = ifelse(p.val < 0.05, as.character(target), '')), hjust = 0, vjust = 0) +
    labs(x = "log2(fold change)", y = "-log10(p value)", color = "Directionality") +
    theme_bw()

```



# Number of significant fold changes, 0% nicotine vs. 3% nicotine vs 6% nicotine, by sex {.tabset}


## Males

```{r}

test.results <- NULL
# make new empty variable that will hold the results

males <- pcr.data %>%
  filter(sex == "m" & (treatment == "ecig0" | treatment == "ecig3" | treatment == "ecig6"))
# filter data to just include males and the treatments that we know

for (i in levels(pcr.data$target)) {
  # initiate loop over the levels of the `target` variable to fit one model per target
  
  data.loop = subset(males, target == i)
  # subset the data to just have one target per loop
  
  data.loop <- data.loop %>% mutate(treatment = factor(treatment, levels = c("ecig0",
                                                                     "ecig3",
                                                                     "ecig6")))
  # relevel the dataset so that the negative control is the reference level. Gets us more interpretable fold change cofficients 
  
  model <- lm(log_fold_change ~ treatment, data = data.loop)
  # fit the model (here the equivalent of a t-test since there are only two levels to the treatment)
  
  anova <- car::Anova(model)
  # fit the anova on the model to get a p-value
  
  # save p value from model for each target
  test.results.list <- data.frame(target = i,
                               p.val = anova$`Pr(>F)`[1],
                               coef.3ecig = coef(model)[[2]],
                               coef.6ecig = coef(model)[[3]])

  test.results <- rbind(test.results, test.results.list)
  # save the results from each loop into the dataframe set up at the beginning to hold them

}

```

### {.tabset}

#### N targets significantly different

```{r}

kable(test.results %>%
        summarise(sum(p.val < 0.05)), col.names = "N targets significantly different (p < 0.05)") %>%
  kableExtra::kable_styling(full_width = FALSE)
# calculate the number of targets with p-values < 0.05

```

#### Significantly different targets (w/ pvalues and average log FCs)

```{r}

kable(test.results %>% filter(p.val < 0.05), col.names = c("target", "p.value", "log fold change (3% ecig)", "log fold change (6% ecig)"), caption = "Significantly different targets", digits = 3) %>%
  kableExtra::kable_styling(full_width = FALSE)
# list the targets falling below the 0.05 threshold


```

#### "Volcano plot"

I hand-made the rt-qPCR equivalent of an RNA-SEQ volcano plot with the natural log fold change plotted against the -log base 10 p-value for each target's control vs. experimental comparison.  
This simply serves as a way to visually represent the tables given above. But they also look cute!  

```{r, fig.width=8, fig.height=12}


test.results <- test.results %>% mutate(color = ifelse(p.val >= 0.05, "Not significant", NA),
                                        color = ifelse(p.val < 0.05 & coef.3ecig < 0, "Decrease", color),
                                        color = ifelse(p.val < 0.05 & coef.3ecig >= 0, "Increase", color),
                                        color = as.factor(color))

three <- ggplot(test.results, aes(x = coef.3ecig, y = -log10(p.val), color = color)) + 
    geom_point() + 
    geom_abline(intercept = -log10(0.05), slope = 0) + 
    geom_label_repel(aes(label = ifelse(p.val < 0.05, as.character(target), '')), hjust = 0, vjust = 0) +
    labs(x = "log2(fold change)", y = "-log10(p value)", color = "Directionality", title = "0% nicotine vs. 3% nicotine") +
    theme_bw()

test.results <- test.results %>% mutate(color = ifelse(p.val >= 0.05, "Not significant", NA),
                                        color = ifelse(p.val < 0.05 & coef.6ecig < 0, "Decrease", color),
                                        color = ifelse(p.val < 0.05 & coef.6ecig >= 0, "Increase", color),
                                        color = as.factor(color))

six <- ggplot(test.results, aes(x = coef.6ecig, y = -log10(p.val), color = color)) + 
    geom_point() + 
    geom_abline(intercept = -log10(0.05), slope = 0) + 
    geom_label_repel(aes(label = ifelse(p.val < 0.05, as.character(target), '')), hjust = 0, vjust = 0) +
    labs(x = "log2(fold change)", y = "-log10(p value)", color = "Directionality", title = "0% nicotine vs. 6% nicotine") +
    theme_bw()

three / six

```


## Females

```{r}

test.results <- NULL

females <- pcr.data %>%
  filter(sex == "f" & (treatment == "ecig0" | treatment == "ecig3" | treatment == "ecig6"))

for (i in levels(pcr.data$target)) {
  
  data.loop = subset(females, target == i)
  
  data.loop <- data.loop %>% mutate(treatment = factor(treatment, levels = c("ecig0",
                                                                               "ecig3",
                                                                               "ecig6")))
  
  model <- lm(log_fold_change ~ treatment, data = data.loop)
  
  anova <- car::Anova(model)
  
  # save p value from model for each target
  test.results.list <- data.frame(target = i,
                               p.val = anova$`Pr(>F)`[1],
                                    coef.3ecig = coef(model)[[2]],
                                    coef.6ecig = coef(model)[[3]])

  test.results <- rbind(test.results, test.results.list)

}

```

### {.tabset}

#### N targets significantly different

```{r}

kable(test.results %>% summarise(sum(p.val < 0.05)), col.names = "N targets significantly different (p < 0.05)") %>%
  kableExtra::kable_styling(full_width = FALSE)

```

#### Significantly different targets (w/ pvalues and average log FCs)

```{r}

kable(test.results %>% filter(p.val < 0.05), col.names = c("target", "p.value", "log fold change (3% ecig)", "log fold change (6% ecig)"), caption = "Significantly different targets", digits = 3) %>%
  kableExtra::kable_styling(full_width = FALSE)

```

#### "Volcano plot"

```{r, fig.width=8, fig.height=12}

test.results <- test.results %>% mutate(color = ifelse(p.val >= 0.05, "Not significant", NA),
                                        color = ifelse(p.val < 0.05 & coef.3ecig < 0, "Decrease", color),
                                        color = ifelse(p.val < 0.05 & coef.3ecig >= 0, "Increase", color),
                                        color = as.factor(color))

three <- ggplot(test.results, aes(x = coef.3ecig, y = -log10(p.val), color = color)) + 
    geom_point() + 
    geom_abline(intercept = -log10(0.05), slope = 0) + 
    geom_label_repel(aes(label = ifelse(p.val < 0.05, as.character(target), '')), hjust = 0, vjust = 0) +
    labs(x = "log2(fold change)", y = "-log10(p value)", color = "Directionality", title = "0% nicotine vs. 3% nicotine") +
    theme_bw()

test.results <- test.results %>% mutate(color = ifelse(p.val >= 0.05, "Not significant", NA),
                                        color = ifelse(p.val < 0.05 & coef.6ecig < 0, "Decrease", color),
                                        color = ifelse(p.val < 0.05 & coef.6ecig >= 0, "Increase", color),
                                        color = as.factor(color))

six <- ggplot(test.results, aes(x = coef.6ecig, y = -log10(p.val), color = color)) + 
    geom_point() + 
    geom_abline(intercept = -log10(0.05), slope = 0) + 
    geom_label_repel(aes(label = ifelse(p.val < 0.05, as.character(target), '')), hjust = 0, vjust = 0) +
    labs(x = "log2(fold change)", y = "-log10(p value)", color = "Directionality", title = "0% nicotine vs. 6% nicotine") +
    theme_bw()

three / six


```


